---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wmms

<!-- badges: start -->
<!-- badges: end -->

Prune a tree to be the most profitable possible subtree that is still connected
to the root node.  If every node has two numeric values, `a` and `b`, the most
profitable subtree will be the one where `sum(a) / sum(b)` over all remaining
nodes is the maximum it could possibly be.  Minimising can be done by
subtracting the `a` values from zero.

This package implements the algorithm in Carlson J., Eppstein D. (2006) The
Weighted Maximum-Mean Subtree and Other Bicriterion Subtree Problems. In: Arge
L., Freivalds R. (eds) Algorithm Theory – SWAT 2006.  SWAT 2006. Lecture Notes
in Computer Science, vol 4059. Springer, Berlin, Heidelberg.

Pure base R is used, yet the package is reasonably quick and has been used to
find optimal trees for every LSOA11 (Lower Layer Super Output Area) in Wales in
less than a minute on a standard laptop.

## Installation

You can install the development version from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("ukgovdatascience/wmms")
```

## Example

Here is an example of building a broadband network out from an existing node.
The cost of extending the network to each node is a function of its distance
from its parent.  We want to minimise the cost of connecting each node, when
averaged over all the nodes that are connected.

In this graph, the root node in the centre has no value, because it already
exists.  The node beside it has the value `10`, because its distance from the
root is `10`.  The `b` value of all nodes is `1`, so that the algorithm will
find an unweighted mean.

```{r ref.label='plot-original', echo = FALSE}
```

Create this graph by making a data frame of edges, one row per edge, with the
columns `from` and `to` referring to node IDs, and `a` and `b` giving the values
of the `to` node.

(One way to obtain such an edgelist from geographical coordinates of each node
is the [`emstreeR`](https://cran.r-project.org/package=emstreeR) package, which
calculates a Euclidean Minimum Spanning Tree.)

```{r create-graph}
library(wmms)

# This graph was invented by Jo Keefe at the UK Department for Digital, Culture,
# Media and Sport.
edges <- data.frame(from = c(1, rep(2:7, each = 5)),
                    to = c(2, 3:32),
                    a = c(10, rep(1:5, 6)),
                    b = 1)
edges
```

```{r plot-original, include = FALSE, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
library(igraph)

nodes <- edges[, c("to", "a", "b")]
nodes <- rbind(nodes, data.frame(to = 1, a = NA, b = NA))

g <- igraph::graph_from_data_frame(edges, vertices = nodes)

plot(g, vertex.label = nodes$a)
```

Now prune the tree to obtain the optimum possible average value over all nodes.
In this example, the values `a` are made negative, so that `a/b` will be
minimised rather than maximised.

```{r maximise}
root_id <- 1

edges$a <- 0 - edges$a

maximum <- weighted_maximum_mean_subtrees(edges, root_id)

maximum$a <- 0 - maximum$a
maximum$value <- 0 - maximum$value

maximum
```

Plot the maximum subtree, and all subtrees connected to it.  The blue nodes are
part of the maximum subtree.  Their average cost is 2.46, which is lower than
the original value of the top node (10).  The cost of appending any other nodes
is higher than 2.46.  The pink nodes comprise another subtree of nodes that have
been averaged together, to achieve the value 2.67, which is lower than the
original value of their top node (5).

```{r plot-optimum, echo = FALSE, fig.width = 8, fig.height = 8, out.width = 600, out.height = 600}
nodes <- maximum[, c("to", "a", "b", "ancestor", "value")]
nodes$value <- sprintf("%.2f", nodes$value)
nodes <- rbind(nodes,
               data.frame(to = 1, a = NA, b = NA, ancestor = NA, value = NA))
nodes$colour = ifelse(nodes$ancestor %in% c(2, 7), nodes$ancestor, NA)

g <- igraph::graph_from_data_frame(maximum, vertices = nodes)

plot(g,
     vertex.label = nodes$value,
     vertex.color = nodes$colour)
```

## Algorithm

A worked example is in a [Google
Sheet](https://docs.google.com/spreadsheets/d/1wCiS0IU6EDkvjXVjRP_MwScVqtBqxqgLj3ODDsErRJ4).

It is also explained in the docs `?algorithm` in the file `R/algorithm.R`.
